System Architecture and Design Specification

### 2025-05-14 nbs/development/architecture.qmd
docs(development): propagate model selection updates to architecture
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

---
title: "System Architecture and Design Specification"
format:
  html:
    mermaid-format: js
    mermaid:
      theme: dark
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
---

@@

**Related Algorithm Specification:** ALG-02 (Stochastic Variational Inference)

@@

```python
def train(self, adata: AnnData, **kwargs) -> Dict[str, Any]:
    # Extract data from AnnData
    data = prepare_anndata(adata)
    u_obs = data["u"]
    s_obs = data["s"]

@@

    return {"losses": losses, "guide": guide}
```

@@

```{mermaid}
graph TB
    Main["PyroVelocityModel"] -->|contains| Dynamics["Dynamics Models"]
    Main -->|contains| Priors["Prior Models"]
    Main -->|contains| Likelihoods["Likelihood Models"]
    Main -->|contains| Observations["Observation Models"]
    Main -->|contains| Guides["Inference Guides"]
    Main -->|contains| Registry["Registry System"]
    Main -->|contains| Factory["Factory System"]

@@

    class Main main;
    class Dynamics,Priors,Likelihoods,Observations,Guides component;
    class DynamicsInterface,PriorsInterface,LikelihoodsInterface,ObservationsInterface,GuidesInterface interface;
    class Registry,Factory system;
```

@@

- **StandardDynamicsModel**: The standard implementation of RNA dynamics
- **LegacyDynamicsModel**: Compatibility implementation matching the legacy model

@@

    DynamicsModel <|.. StandardDynamicsModel : implements
    DynamicsModel <|.. LegacyDynamicsModel : implements
```

@@

>>>>

---
title: "System Architecture and Design Specification"
format:
  html:
    mermaid-format: js
    mermaid:
      theme: dark
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
bibliography: ../references.bib
csl: ../bibstyle.csl
---

@@

**Related Algorithm Specification:** ALG-03 (Stochastic Variational Inference)

@@

```python
def train_svi(self, adata: AnnData, **kwargs) -> Dict[str, Any]:
    # Extract data from AnnData
    data = prepare_anndata(adata)
    u_obs = data["u"]
    s_obs = data["s"]

@@

    return {"losses": losses, "guide": guide, "posterior_samples": guide.get_posterior()}
```

**Algorithm ID:** ALG-IMP-02

**Algorithm Name:** Markov Chain Monte Carlo for RNA Velocity

**Related Algorithm Specification:** ALG-04 (Markov Chain Monte Carlo Inference)

**Implementation Approach:** NumPyro MCMC with NUTS

```python
def train_mcmc(self, adata: AnnData, **kwargs) -> Dict[str, Any]:
    # Extract data from AnnData
    data = prepare_anndata(adata)
    u_obs = data["u"]
    s_obs = data["s"]

    # Set up MCMC
    num_samples = kwargs.get("num_samples", 1000)
    num_warmup = kwargs.get("num_warmup", 500)
    num_chains = kwargs.get("num_chains", 4)

    # Create MCMC kernel
    kernel = numpyro.infer.NUTS(self.forward)
    mcmc = numpyro.infer.MCMC(
        kernel,
        num_samples=num_samples,
        num_warmup=num_warmup,
        num_chains=num_chains,
        progress_bar=kwargs.get("progress_bar", True)
    )

    # Run MCMC
    mcmc.run(jax.random.PRNGKey(0), u_obs=u_obs, s_obs=s_obs)

    # Get diagnostics
    diagnostics = {
        "r_hat": mcmc.get_diagnostics()["r_hat"],
        "effective_sample_size": mcmc.get_diagnostics()["effective_sample_size"],
        "acceptance_rate": mcmc.get_diagnostics()["acceptance_rate"],
        "divergences": mcmc.get_diagnostics()["divergences"]
    }

    # Get posterior samples
    posterior_samples = mcmc.get_samples()

    return {
        "posterior_samples": posterior_samples,
        "diagnostics": diagnostics,
        "mcmc": mcmc
    }
```

@@

```{mermaid}
graph TB
    Main["PyroVelocityModel"] -->|contains| Dynamics["Dynamics Models"]
    Main -->|contains| Priors["Prior Models"]
    Main -->|contains| Likelihoods["Likelihood Models"]
    Main -->|contains| Observations["Observation Models"]
    Main -->|contains| Guides["Inference Guides"]
    Main -->|contains| MCMC["MCMC Models"]
    Main -->|contains| Registry["Registry System"]
    Main -->|contains| Factory["Factory System"]
    Main -->|contains| Validation["Validation Framework"]

    Dynamics -->|implements| DynamicsInterface["DynamicsModel Protocol"]
    Priors -->|implements| PriorsInterface["PriorModel Protocol"]
    Likelihoods -->|implements| LikelihoodsInterface["LikelihoodModel Protocol"]
    Observations -->|implements| ObservationsInterface["ObservationModel Protocol"]
    Guides -->|implements| GuidesInterface["InferenceGuide Protocol"]
    MCMC -->|implements| MCMCInterface["MCMCModel Protocol"]

@@

    class Main main;
    class Dynamics,Priors,Likelihoods,Observations,Guides,MCMC component;
    class DynamicsInterface,PriorsInterface,LikelihoodsInterface,ObservationsInterface,GuidesInterface,MCMCInterface interface;
    class Registry,Factory system;
    class Validation,ParameterRecovery,ModelComparison,ValidationMetrics validation;
```

@@

- **StandardDynamicsModel**: The standard implementation of RNA dynamics
- **LegacyDynamicsModel**: Compatibility implementation matching the legacy model
- **NonlinearDynamicsModel**: Implementation supporting nonlinear dynamics with numerical integration

@@

    class NonlinearDynamicsModel {
        +name: str
        +description: str
        +shared_time: bool
        +t_scale_on: bool
        +ode_solver: ODESolver
        +forward(context) Dict
        +steady_state(alpha, beta, gamma) Tuple
        +solve_ode(initial_state, params, times) Tensor
    }

    DynamicsModel <|.. StandardDynamicsModel : implements
    DynamicsModel <|.. LegacyDynamicsModel : implements
    DynamicsModel <|.. NonlinearDynamicsModel : implements
```

@@

#### MCMC Models {#sec-mcmc-models}

MCMC models implement Markov Chain Monte Carlo sampling for Bayesian inference.

- **NUTSModel**: Implements the No U-Turn Sampler for MCMC
- **HMCModel**: Implements Hamiltonian Monte Carlo for MCMC

```{mermaid}
classDiagram
    class MCMCModel {
        <<Protocol>>
        +sample(model, data, num_samples, num_chains) Dict
        +get_diagnostics() Dict
    }

    class NUTSModel {
        +name: str
        +description: str
        +sample(model, data, num_samples, num_chains) Dict
        +get_diagnostics() Dict
        +compute_r_hat() Tensor
        +compute_effective_sample_size() Tensor
    }

    class HMCModel {
        +name: str
        +description: str
        +sample(model, data, num_samples, num_chains) Dict
        +get_diagnostics() Dict
        +compute_r_hat() Tensor
        +compute_effective_sample_size() Tensor
    }

    MCMCModel <|.. NUTSModel : implements
    MCMCModel <|.. HMCModel : implements
```

#### Validation Framework {#sec-validation-framework}

The validation framework provides tools for model validation and comparison.

- **ParameterRecoveryValidator**: Validates model's ability to recover known parameters from synthetic data
- **ModelComparisonFramework**: Compares different model implementations and configurations
- **ValidationMetrics**: Computes metrics for quantifying validation performance

```{mermaid}
classDiagram
    class ParameterRecoveryValidator {
        <<Protocol>>
        +generate_synthetic_data(model_config, prior_config, num_cells, num_genes) Dict
        +validate_parameter_recovery(model_config, synthetic_data) Dict
    }

    class ModelComparisonFramework {
        <<Protocol>>
        +compare_models(models, data, metrics) Dict
        +rank_models(comparison_results, criteria) Dict
    }

    class ValidationMetrics {
        <<Protocol>>
        +compute_parameter_metrics(true_params, estimated_params) Dict
        +compute_coverage_metrics(true_params, credible_intervals) Dict
        +compute_model_comparison_metrics(models, data) Dict
    }

    class StandardParameterRecoveryValidator {
        +generate_synthetic_data(model_config, prior_config, num_cells, num_genes) Dict
        +validate_parameter_recovery(model_config, synthetic_data) Dict
        +compute_validation_metrics(true_params, estimated_params) Dict
    }

    class StandardModelComparisonFramework {
        +compare_models(models, data, metrics) Dict
        +rank_models(comparison_results, criteria) Dict
        +visualize_comparison(comparison_results) Dict
    }

    ParameterRecoveryValidator <|.. StandardParameterRecoveryValidator : implements
    ModelComparisonFramework <|.. StandardModelComparisonFramework : implements
```
